Serum Biomarkers of Nutrition Pre and Post CFTR Modulator Use in Children with Cystic Fibrosis

### 2022-04-06 Edith Zemanick/CF Nutrition/cf_nutrition.Rmd
basic comparisons done

```{r data cleaning}
# Import
df = read.csv("~/Dropbox/Work/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv")
# Only those on effective modulators
on_effective = unique(df$Patient.ID[which(df[,'Modulator'] %in% c('Kalydeco','Trikafta'))])
df = df %>% filter(Patient.ID %in% on_effective)
# Non-highy efective to "None
df$Modulator[!df$Modulator %in% c('Kalydeco','Trikafta')] = "None"
# Before or after
df$Before_After = df$Age.at.Test..Years. < df$Age.at.Start..years.
df$Before_After = factor(df$Before_After,levels = c(FALSE,TRUE),labels = c("After","Before"))
# Count number of measures 
num_values = df %>% group_by(Patient.ID,Before_After) %>% 
  summarise(n = sum(!is.na(X25OH.Vitamin.D)),.groups = "drop") %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = n)
num_values$`NA` = NULL
colnames(num_values)[2:3] = c("vit_measures_after","vit_measures_before")
df = left_join(df,num_values,by = "Patient.ID")
# At least two annual nutritional measurements prior to first modulator start date
two_measures = num_values$Patient.ID[which(num_values$`vit_measures_before` >= 2)]
# At least one measurement of nutritional markers and one weight/height measurement >= three months post-modulator
time_after = df %>% 
  filter(Age.at.Test..Years. >= (Age.at.Start..years. + 0.25),!is.na(X25OH.Vitamin.D)) %>%
  pull(Patient.ID) %>% unique(.)
# Filter out those without above criteria, get first modulator
df = df %>% filter(Patient.ID %in% intersect(two_measures,time_after)) %>%
  group_by(Patient.ID) %>% 
  mutate(first_mod = head(Modulator[Modulator != "None"],1)) %>% ungroup()
# Genotype group
geno = read_excel("/Users/timvigers/Dropbox/Work/CF/CF Info/Sweat Tests from Internal Database_genotype class.xlsx")
geno = geno[,c('PatientID','genoRisk')]
df = left_join(df,geno,by = c('Patient.ID' = 'PatientID'))
# Time prior to first modulator
years = df %>% mutate(years_to_start = Age.at.Test..Years. - Age.at.Start..years.) %>%
  group_by(Patient.ID,Before_After) %>% 
  mutate(years_to_start_max = max(abs(years_to_start))) %>% ungroup() %>%
  select(Patient.ID,Before_After,years_to_start_max) %>% distinct(.) %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = years_to_start_max)
years$`NA` = NULL
colnames(years)[2:3] = c('max_years_to_start','max_years_from_start')
df = left_join(df,years,by = "Patient.ID")
```

@@

```{r results='asis'}

```

>>>>

```{r data cleaning}
# Import
df = read.csv("~/Dropbox/Work/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv")
# Only those on effective modulators
on_effective = unique(df$Patient.ID[which(df[,'Modulator'] %in% c('Kalydeco','Trikafta'))])
df = df %>% filter(Patient.ID %in% on_effective)
# Non-highy efective to "None
df$Modulator[!df$Modulator %in% c('Kalydeco','Trikafta')] = "None"
# Before or after
df$Before_After = df$Age.at.Test..Years. < df$Age.at.Start..years.
df$Before_After = factor(df$Before_After,levels = c(FALSE,TRUE),labels = c("After","Before"))
# Count number of measures 
num_values = df %>% group_by(Patient.ID,Before_After) %>% 
  summarise(n = sum(!is.na(X25OH.Vitamin.D)),.groups = "drop") %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = n)
num_values$`NA` = NULL
colnames(num_values)[2:3] = c("vit_measures_after","vit_measures_before")
df = left_join(df,num_values,by = "Patient.ID")
# At least two annual nutritional measurements prior to first modulator start date
two_measures = num_values$Patient.ID[which(num_values$`vit_measures_before` >= 2)]
# At least one measurement of nutritional markers and one weight/height measurement >= three months post-modulator
time_after = df %>% 
  filter(Age.at.Test..Years. >= (Age.at.Start..years. + 0.25),!is.na(X25OH.Vitamin.D)) %>%
  pull(Patient.ID) %>% unique(.)
# Filter out those without above criteria, get first modulator
df = df %>% filter(Patient.ID %in% intersect(two_measures,time_after)) %>%
  group_by(Patient.ID) %>% 
  mutate(first_mod = head(Modulator[Modulator != "None"],1)) %>% ungroup()
# Genotype group
geno = read_excel("/Users/timvigers/Dropbox/Work/CF/CF Info/Sweat Tests from Internal Database_genotype class.xlsx")
geno = geno[,c('PatientID','genoRisk')]
df = left_join(df,geno,by = c('Patient.ID' = 'PatientID'))
# Time prior to first modulator
years = df %>% group_by(Patient.ID,Before_After) %>% 
  mutate(years_to_start_max = max(abs(Age.at.Test..Years. - Age.at.Start..years.))) %>% 
  select(Patient.ID,Before_After,years_to_start_max) %>% 
  ungroup() %>% distinct(.) %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = years_to_start_max)
years$`NA` = NULL
colnames(years)[2:3] = c('max_years_to_start','max_years_from_start')
df = left_join(df,years,by = "Patient.ID")
df$years = df$Age.at.Test..Years. - df$Age.at.Start..years.
```

@@

For each participant, we selected the pre and post visit closest to modulator start (and at least 3 months after start) with both a vitamin A and FEV1 % predicted value measure. Some visits were on different days and had different measures, so some missing values were filled in with the closest value (within 1 month).

```{r results='asis'}
nutrition_vars = c("VitaminA_Retinol","X25OH.Vitamin.D","Vitamin.E.Alpha",
                   "BMI","FEV1...pred")
# Get the most recent 
paired_df = df %>% 
  select(Patient.ID,Before_After,years,first_mod,all_of(nutrition_vars)) %>% 
  filter(years >= 0.25 | years < 0) %>%
  mutate(years = round(years)) %>%
  group_by(Patient.ID,years) %>%
  fill(all_of(nutrition_vars),.direction = "downup") %>%
  drop_na(any_of(c("VitaminA_Retinol","FEV1...pred"))) %>%
  group_by(Patient.ID,Before_After) %>%
  slice_min(abs(years)) %>% distinct() %>% filter(row_number()==1) 
paired_df$Before_After = relevel(paired_df$Before_After,ref = "Before")

f = as.formula(paste0("Before_After ~ ",paste0(nutrition_vars,collapse = "+")))
p = paired(f,data = paired_df,id = Patient.ID,strata = first_mod)
summary(p,
        labelTranslations = list(VitaminA_Retinol = "Vitamin A Retinol",
                                 X25OH.Vitamin.D = "25OH-Vitamin D",
                                 Vitamin.E.Alpha = "FEV1 % Predicted",
                                 first_mod = "First Modulator"))
```

# Piecewise longitudinal models

```{r}
# Set up data again
df = read.csv("~/Dropbox/Work/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv")
df$years = df$Age.at.Test..Years. - df$Age.at.Start..years.
t = df %>% group_by(Patient.ID) %>% summarise(n = sum(!is.na(VitaminA_Retinol)))
```

```{r}
outcome = "VitaminA_Retinol"
piecewise_mod = function(outcome,data){
  data = data %>% select(Patient.ID,years,all_of(outcome)) %>% drop_na()
  plot = ggplot(data,aes_string(x = "years",y = outcome,group = "Patient.ID")) + 
    geom_point() + geom_line()
}
```
